Appendix S4. The interpretation of the estimated trait change effect size coefficient

### 2024-11-30 manuscript/Appendix_S4_v01.Rmd

---
title: "Appendix S4. The relative importance of evolution in populations and communities"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

@@

# Logistic growth and trait evolution

We can create simulations of evolutionary rescue in a population, where adaptive evolution rescues a population from extinction, at a speed that depends on the heritability level in the system. Population dynamics are modelled as:

$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \alpha_{ij} N_{j,t}} $$

where \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, and $h^2$ is the heritability of the trait *x*. For the simulation we use \(W_{max}\) = 2, *P* = 1, and *w* = 2.

The change in the average trait value each time step is given by:

$$ d_{i,t+1} = k d_{i,t} $$
where \( k = \frac{w + (1-h^2) P}{w + P} \) and \( d_{i,t} = E_t - x_{i,t} \).

## Population dynamics simulation

```{r}
### 1 species, h2=0 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- N
gdat$time <- 1:t
gdat$h2 <- h2
gdat$x1 <- x$x1
gdat$E <- E
```
```{r}
### 1 species, h2=0.01 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.01
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.02 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.02
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.03 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.03
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.04 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.04
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.05 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.05
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.075 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.075
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.1 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.1
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.4 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.4
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=0.9 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.9
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r}
### 1 species, h2=1 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 1
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```

```{r}
### Plot of time series ###
gdat$h2 <- as.factor(gdat$h2)
ggplot2::ggplot(gdat, aes(time, N1, col=h2)) + geom_point()
### ###
```

## HMSC model fit

The goal of fitting the data to a statistical model is to estimate the relative important of trait evolution for the population dynamics.

```{r}
### Model 1, h2=00 ###
dat <- cbind(log(gdat$N1[gdat$h2==0]),gdat$x1[gdat$h2==0])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.1.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.1.sample <- sampleMcmc(m.1.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m1.post.hmsc <- convertToCodaObject(m.1.sample)
```

```{r warning=FALSE}
VP1 <- computeVariancePartitioning(m.1.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.1.sample,VP1,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.1.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

```{r}
### Model 2, h2=0.01 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.01]),gdat$x1[gdat$h2==0.01])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.2.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m2.post.hmsc <- convertToCodaObject(m.2.sample)
```

```{r warning=FALSE}
VP2 <- computeVariancePartitioning(m.2.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.2.sample,VP2,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.2.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

```{r}
### Model 3, h2=0.02 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.02]),gdat$x1[gdat$h2==0.02])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.3.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.3.sample <- sampleMcmc(m.3.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m3.post.hmsc <- convertToCodaObject(m.3.sample)
```

```{r warning=FALSE}
VP3 <- computeVariancePartitioning(m.3.sample,group=c(1,1,1,2),groupnames=c("Env","x1"))
plotVariancePartitioning(m.3.sample,VP3,cols=c("white","skyblue","darkgrey"),args.legend=list(cex=0.75,bg="transparent"))
```
```{r}
# prepare for plot
pred2 <- predict(m.3.sample)
hold1 <- list()
hold2 <- list()
for(i in 1:length(N0)){
  hold1[[i]] <- as.numeric(Y[,i])
  data <- lapply(pred2, function(x) x[, colnames(x)[i], drop = FALSE])
  hold2[[i]] <- matrix(unlist(data), nrow=length(pred2), byrow=TRUE)
}
y <- unlist(hold1) # vector of each species' observed N across time points
yrep <- as.matrix(data.frame(hold2))
### Mod2 plot ###
par(mgp=c(2,0.45,0), tcl=-0.4, mar=c(1.3,1.2,0,0))
color_scheme_set("brightblue")
#ppc_dens_overlay(y, yrep[1:50, ])
ppc_intervals(y,yrep,x=dat$time[1:(t-1)],prob = 0.95) +
  labs(x = "time",y = "ln(N)",) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  theme(legend.position = "none")
```

```{r}
### Model 4, h2=0.03 ###
dat <- cbind(log(gdat$N1[gdat$h2==0.03]),gdat$x1[gdat$h2==0.03])
dat <- as.data.frame(dat)
colnames(dat) <- c("N1","x1")
dat$time <- 1:t
dat <- as.data.frame(dat)
df <- data.frame(dat[(2:t),-3])
colnames(df) <- c("Nt1", "xt1")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t - 1)])
# prepare data in HMSC format
Y <- as.matrix(df$Nt1)
XData <- data.frame(cbind(E[1:(t-1)],E[1:(t-1)]^2,abs(dat$x1[2:t] - dat$x1[1:(t-1)])))
colnames(XData) <- c("E","Esq","dx1")

studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)

m.4.hmsc = Hmsc(Y = Y, XData = XData, XFormula = ~E+Esq+dx1, studyDesign = studyDesign, ranLevels = list(sample = rL))
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 1000*thin
verbose <- 500*thin
# sample MCMC
m.4.sample <- sampleMcmc(m.4.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
m4.post.hmsc <- convertToCodaObject(m.4.sample)
```

>>>>

---
title: "Appendix S4. The interpretation of the estimated trait change effect size coefficient"
author:
- Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
- Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

@@

# Empirical data

To correctly interpret correlation of trait change with population density of the species (as seen in Fig. 3) we first need to check the raw data.

```{r fig1, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Figure S05 1"), echo=FALSE}
knitr::include_graphics("../output/fig_S05_1.pdf")
```

From the empirical data we can detect that species Rn is drastically increasing in frond size over the course of the experiment, with a peak in trait change during week 5 and 7. Simultaneously it is only slowly increasing in population density, with slightly lower density during week 5. This leads to the highest value in $|\Delta x|$ during a lower value in the population density of species Rn. Similarly species Wc had large trait changes, during time points when species Lm and Sp showed an increase in population density. HMSC draws a linear correlation between both and in (\@ref(fig:fig2)) it can be seen how HMSC infers these correlations (the logarithmic scale of $|\Delta x|$ was used for the figure).

```{r fig2, out.width="90%", include=TRUE, fig.align="center", fig.cap=c("Figure S05 2"), echo=FALSE}
knitr::include_graphics("../output/fig_S05_2.pdf")
```

It should be noted that species Rn was the only to strongly increase in frond size in this experiment, which may the result from its decrease in population size. Species Wc decreased in frond size, thus occupying less space on the water surface and potentially freeing space for the other species to occupy. We speculate on this, but the manuscript for the original data will have more information.
